Multimodal Information Extraction of Supermarket Leaflets

### 2025-02-26 paper/main.tex
Add final changes.

% !TEX root = main_ilyi.tex

@@

@@

\subsection{Experiments}
Evaluating the performance of instance segmentation models requires an analysis of how well detected objects align with ground-truth annotations. The primary metric used in this assessment is \emph{Mean Average Precision (mAP)}, which quantifies detection accuracy at different levels of overlap between predicted and actual objects. The core component of mAP is Intersection over Union (IoU). IoU is defined as the ratio of the intersection area between the predicted and ground-truth segmentation masks to their union. A higher IoU indicates a better match between the prediction and the actual object, while a lower IoU suggests that the detected region is misaligned or includes excessive background. In the case of bounding boxes, IoU is calculated based on the overlap of rectangular regions, whereas for instance segmentation, it is computed on a pixel level, capturing the true shape of the detected objects. To evaluate model performance, different IoU thresholds are applied. A commonly used metric is mAP@50, which considers a detection correct if the IoU between the predicted and ground-truth mask exceeds 50\%. This threshold provides insight into how well a model identifies objects in general but does not strictly enforce precision in localization. A more comprehensive metric is mAP@50-95, which averages precision across multiple IoU thresholds, ranging from 50\% to 95\% in increments of 5\%. This approach penalizes models that fail to produce accurate segmentations and is particularly useful for assessing how well the model distinguishes closely packed or irregularly shaped deals.

@@

The first set of experiments focused on comparing different YOLO variants trained on the dataset. Each model was trained for 20 epochs without additional data augmentation. The results, shown in Figure \ref{fig:all_yolo_results} (a), present the mAP50 (mask) scores for YOLOv8 and YOLOv11 in their small, medium, and large configurations. The findings indicate that the smallest version of YOLOv8 achieved the highest validation mAP50, outperforming larger models in this setting. A second analysis, visualized in Figure \ref{fig:all_yolo_results} (b), examines mAP50-95 scores for the same models, reinforcing the observation that a lightweight YOLO model is sufficient for this particular task. The lower performance of larger models suggests that the dataset’s structure, with relatively uniform layouts, does not require highly complex architectures to achieve strong results.

@@

Following this baseline evaluation, additional experiments investigated the impact of data augmentation on segmentation performance. Various augmentation strategies were applied, including translation, scaling, horizontal flipping, and color transformations. The goal was to determine whether these transformations improved the model’s ability to generalize across different leaflet layouts. Each augmented model was trained for ten additional epochs, and the results are displayed in Figure \ref{fig:yolo8_results}. While certain augmentation techniques led to slight improvements in segmentation accuracy, particularly for models trained with random translations, the overall best-performing model remained the one trained without any augmentation. This outcome suggests that the dataset already contains sufficient variation to ensure robust generalization without the need for extensive artificial augmentations.

@@

\subsubsection{Traditional and Hybrid Approaches for Supermarket Leaflet Information Extraction}  
Information extraction from supermarket leaflets has traditionally been approached through a pipeline of sequential sub-tasks, each addressed by specialized systems. These methods, while extensively studied and methodologically diverse, are typically divided into three main stages: text detection, text recognition, and key-value pair extraction. Although effective in certain scenarios, these approaches suffer from inherent limitations, including computational overhead, error propagation, and suboptimal performance on edge cases. Recent advancements in LLMs have led to the development of hybrid OCR-LLM architectures, which combine the precision of OCR with the contextual reasoning capabilities of LLMs. Below, we formalize these approaches and discuss their strengths and limitations.

@@

\subsection{Dataset Creation}

@@

Even though the Leaflet-IE dataset is relatively small, the reader will be able to see that the dataset is sufficient to evaluate the performance of different IE appraoches as well as to train an competitive end-to-end model. However, as the reader may notice, the dataset is solely focused on single product deals at this point and explicitly does not include
% TODO: Missing Text (ILYI)

\begin{figure*}[h!]
\centering
\includegraphics[width=0.5\linewidth]{figures/ie_samples.png}
\caption{Sample images with annotations from the custom dataset.}
\label{fig:ie_dataset_samples}
\end{figure*}

@@

% TODO: Remove cause of ambiguity? (ILYI)
% The experiments aim to identify the most effective methodology for extracting structured information from supermarket leaflets by evaluating three candidate approaches: (1) Hybrid OCR + LLM pipelines, (2) end-to-end Vision-Encoder-Decoder models, and (3) LVLMs. The evaluation is structured into four sequential phases to systematically assess performance and scalability.

First, Plain Deal OCR Performance establishes a baseline by measuring the accuracy of standalone OCR systems in extracting textual content from leaflets, highlighting challenges such as font variability and noisy backgrounds. Second, OCR + LLM Performance evaluates the combined pipeline, where OCR outputs are processed by large language models to resolve ambiguities and extract structured information. Third, Donut Fine-Tuning Performance assesses the Vision-Encoder-Decoder model's effectiveness after fine-tuning on leaflet-specific data, focusing on its ability to handle multimodal inputs without intermediate OCR steps. Finally, the MAIN experiment conducts a comprehensive comparison of Hybrid OCR+LLM, Donut, and LVLM approaches, analyzing their accuracy, robustness, and computational efficiency in real-world leaflet data scenarios.

\subsubsection{Implementation Details}
The experiments were conducted on two distinct computational workstations - station 1 equipped with an NVIDIA RTX 4080 GPU and station 2 with an NVIDIA GTX 1080 GPU. Therefore, the experiments could be conducted in parallel, ensuring efficient model training and evaluation.

The software stack includegraphics Python with PyTorch\footnote{\href{https://pytorch.org/}{https://pytorch.org/}} as the primary deep learning framework, HuggingFace Transformers \footnote{\href{https://huggingface.co/transformers/}{https://huggingface.co/transformers/}}
for leveraging pre-trained models, ollama \footnote{\href{https://ollama.com/}{https://ollama.com/}}
and lmstudio \footnote{\href{https://lmstudio.ai/}{https://lmstudio.ai/}} for LVLMs, and OpenCV \footnote{\href{https://opencv.org/}{https://opencv.org/}} for image preprocessing and data augmentation, as well as various other libraries for data pre-and post-processing, visualization, data management and monitoring \footnote{The codebase is available at \href{https://github.com/ilyii/leaflets}{https://github.com/ilyii/leaflets}}.
% TODO: Remove this section or move it to the appendix or introduction because it is also relevant for deal detection chapter? (ILYI)

@@

\indent{\textbf{Results}.} Figure \ref{fig:eval_ocr_accuracies} presents a comparative analysis of various OCR models and normalization levels based on per-entity accuracy. The results highlight several key trends regarding the recognition performance of different entities.

@@

To avoid confusion with the normalization used in the OCR experiment, the OCR results were kept raw, as normalization may remove valuable information. The LLMs are expected to handle such inconsistencies and errors in the input. Instead, here the entity values were evaluated in both raw form and with a normalization applied for the comparability between the LLM prediction and the ground truth value. The normalization procedure includes usual character replacements and filtering based on linguistic artifacts commonly encountered in OCR errors. The specific normalization procedure can be found in \secref{app:ocr_llm_normalization}.
\paragraph{Results:}
The entity-averaged accuracies and Levenshtein distances for the raw OCR output are depicted in Figures \ref{fig:eval_ocr_llm_accuracies_avg} and \ref{fig:eval_ocr_llm_levdist_avg}, respectively.

@@

\subsection{Donut Fine-Tuning Performance}
The experiment aims to evaluate the fine-tuning performance of the Vision Encoder-Decoder model, Donut, pretrained on the CORD dataset, to determine its capability in extracting end-to-end information from supermarket leaflet deal images. Conducted on an NVIDIA RTX 4080 (Station 1), the model has a vocabulary size of 57,580 and 201M parameters. The training configuration includes 10 epochs, validation performed every epoch, gradient clipping at 2.0, 800 training samples per epoch, a learning rate of 8e-6, batch sizes of 16 for both training and validation, 400 warmup steps, and 16-bit mixed precision.

@@

\subsection{Comparison of Hybrid OCR+LLM, Donut, and LVLM}

This section presents a comparative evaluation of a hybrid OCR+LLM pipeline against end-to-end models, specifically Donut and LVLMs, for IE from supermarket deals. The assessment focuses on model performance using the best OCR+LLM configuration and two end-to-end approaches.

\subsubsection{Experimental Setup}
% TODO: Remove \subsection{Comparison of Hybrid OCR+LLM, Donut, and LVLM} because it is the same as the following? (ILYI)
The evaluation framework was designed to compare the performance of three distinct model categories: OCR+LLM pipelines, end-to-end Vision Encoder-Decoder models, and LVLMs. For the OCR+LLM approach, DocTR served as the OCR engine, paired with Qwen 2.5-7B and LLaMA 3.1-8B as the large language models, selected for their comparable performance. The Donut model, an end-to-end Vision Encoder-Decoder architecture optimized for document understanding, was included to assess its capabilities in handling structured document data. Additionally, MiniCPM-V \cite{yao2024} and LLaMA 3.2-Vision \cite{touvron2023} were chosen as representative LVLMs to evaluate their multimodal reasoning abilities.

@@

\subsection{Database}
\begin{figure*}[h!]
    \centering
    \includegraphics[width=0.8\linewidth]{figures/application/db_uml.png}
    \caption{Database schema for storing metadata for supermarket leaflets and extracted deals.}
    \label{fig:application_db_uml}
\end{figure*}
For storing the extracted deals, a lightweight and efficient solution was required. SQLite was chosen due to its simplicity and fast performance, making it well-suited for handling the structured but relatively small-scale dataset. After several weeks of scraping supermarket leaflets, the dataset grew to include 540 leaflets from 20 different supermarkets. The database is structured into three primary tables (cf. Figure \ref{fig:application_db_uml}): one for supermarket metadata, one for storing the leaflets and one containing the extracted deals. The total number of extracted deals in the database adds up to 47,829. Among these, the vast majority (46,869) are classified as standard deals, while 607 fall under product category deals, 144 correspond to coupons and 208 belong to miscellaneous categories. Since the structure and content of non-standard deals often differ significantly from regular deals featuring loyalty points, bulk discounts, or percentage-based reductions across entire product groups, the primary focus remains on standard deals. By structuring the data efficiently, the database serves as a robust backend for the subsequent web application, facilitating fast queries and scalable deal retrieval.

@@

Building on the database, a web-based application was developed to provide users with an interactive way to explore supermarket deals, which is shown in Figure \ref{fig:application_web_overview}. The core functionality centers around an intuitive search system that allows users to find deals based on specific product names. Additionally the interface supports filtering options, enabling users to refine searches by supermarket chain or deal validity period. Each deal is displayed with its corresponding image snippet, extracted metadata, and pricing information, making it easy to compare offers across different retailers.

@@

>>>>

% !TEX root = main.tex

@@

% Table of contents
% \tableofcontents

@@

\subsection{Experiments}
Evaluating the performance of instance segmentation models requires an analysis of how well detected objects align with ground-truth annotations. The primary metric used in this assessment is \emph{Mean Average Precision (mAP)}, which quantifies detection accuracy at different levels of overlap between predicted and actual objects. The core component of mAP is Intersection over Union (IoU). IoU is defined as the ratio of the intersection area between the predicted and ground-truth segmentation masks to their union. A higher IoU indicates a better match between the prediction and the actual object, while a lower IoU suggests that the detected region is misaligned or includes excessive background. In the case of bounding boxes, IoU is calculated based on the overlap of rectangular regions, whereas for instance segmentation, it is computed on a pixel level, capturing the true shape of the detected objects. To evaluate model performance, different IoU thresholds are applied. A commonly used metric is mAP@50, which considers a detection correct if the IoU between the predicted and ground-truth mask exceeds 50\%. This threshold provides insight into how well a model identifies objects in general but does not strictly enforce precision in localization. A more comprehensive metric is mAP@50-95, which averages precision across multiple IoU thresholds, ranging from 50\% to 95\% in increments of 5\%. This approach penalizes models that fail to produce accurate segmentations and is particularly useful for assessing how well the model distinguishes closely packed or irregularly shaped deals. The implemention details can be found in \secref{app:implementation_details}.

@@

The first set of experiments focused on comparing different YOLO variants trained on the dataset. Each model was trained for 20 epochs without additional data augmentation. The results, shown in \figref{fig:all_yolo_results} (a), present the mAP50 (mask) scores for YOLOv8 and YOLOv11 in their small, medium, and large configurations. The findings indicate that the smallest version of YOLOv8 achieved the highest validation mAP50, outperforming larger models in this setting. A second analysis, visualized in \figref{fig:all_yolo_results} (b), examines mAP50-95 scores for the same models, reinforcing the observation that a lightweight YOLO model is sufficient for this particular task. The lower performance of larger models suggests that the dataset’s structure, with relatively uniform layouts, does not require highly complex architectures to achieve strong results.

@@

Following this baseline evaluation, additional experiments investigated the impact of data augmentation on segmentation performance. Various augmentation strategies were applied, including translation, scaling, horizontal flipping, and color transformations. The goal was to determine whether these transformations improved the model’s ability to generalize across different leaflet layouts. Each augmented model was trained for ten additional epochs, and the results are displayed in \figref{fig:yolo8_results}. While certain augmentation techniques led to slight improvements in segmentation accuracy, particularly for models trained with random translations, the overall best-performing model remained the one trained without any augmentation. This outcome suggests that the dataset already contains sufficient variation to ensure robust generalization without the need for extensive artificial augmentations.

@@

\subsubsection{Traditional and Hybrid Approaches}  
Information extraction from supermarket leaflets has traditionally been approached through a pipeline of sequential sub-tasks, each addressed by specialized systems. These methods, while extensively studied and methodologically diverse, are typically divided into three main stages: text detection, text recognition, and key-value pair extraction. Although effective in certain scenarios, these approaches suffer from inherent limitations, including computational overhead, error propagation, and suboptimal performance on edge cases. Recent advancements in LLMs have led to the development of hybrid OCR-LLM architectures, which combine the precision of OCR with the contextual reasoning capabilities of LLMs. Below, we formalize these approaches and discuss their strengths and limitations.

@@

\subsection{Dataset: Leaflet-IE}

@@

Even though the Leaflet-IE dataset is relatively small, the reader will be able to see that the dataset is sufficient to evaluate the performance of different IE appraoches as well as to train an competitive end-to-end model. However, as the reader may notice, the dataset is solely focused on single product deals at this point and explicitly does not include more complex deals like "buy-one-get-one-free" or "buy-two-for-the-price-of-one" deals. In this stage, the dataset is designed to be a starting point for further research and development, and the inclusion of more complex deals is planned for future work.
\begin{figure*}[h!]
\centering
\includegraphics[width=0.5\linewidth]{figures/ie_samples.png}
\caption{Sample images with annotations from the custom dataset.}
\label{fig:ie_dataset_samples}
\end{figure*}

@@

First, Plain Deal OCR Performance establishes a baseline by measuring the accuracy of standalone OCR systems in extracting textual content from leaflets, highlighting challenges such as font variability and noisy backgrounds. Second, OCR + LLM Performance evaluates the combined pipeline, where OCR outputs are processed by large language models to resolve ambiguities and extract structured information. Third, Donut Fine-Tuning Performance assesses the Vision-Encoder-Decoder model's effectiveness after fine-tuning on leaflet-specific data, focusing on its ability to handle multimodal inputs without intermediate OCR steps. Finally, the MAIN experiment conducts a comprehensive comparison of Hybrid OCR+LLM, Donut, and LVLM approaches, analyzing their accuracy, robustness, and computational efficiency in real-world leaflet data scenarios. The hardware and software related implementation details can be found in \secref{app:implementation_details}.

@@

\indent{\textbf{Results}.} \figref{fig:eval_ocr_accuracies} presents a comparative analysis of various OCR models and normalization levels based on per-entity accuracy. The results highlight several key trends regarding the recognition performance of different entities.

@@

To avoid confusion with the normalization used in the OCR experiment, the OCR results were kept raw, as normalization may remove valuable information. The LLMs are expected to handle such inconsistencies and errors in the input. Instead, here the entity values were evaluated in both raw form and with a normalization applied for the comparability between the LLM prediction and the ground truth value. The normalization procedure includes usual character replacements and filtering based on linguistic artifacts commonly encountered in OCR errors. The specific normalization procedure can be found in \secref{app:ocr_llm_normalization}.
\paragraph{Results:}
The entity-averaged accuracies and Levenshtein distances for the raw OCR output are depicted in  \figref{fig:eval_ocr_llm_accuracies_avg} and \figref{fig:eval_ocr_llm_levdist_avg}, respectively.

@@

\subsubsection{Donut Fine-Tuning Performance}
The experiment aims to evaluate the fine-tuning performance of the Vision Encoder-Decoder model, Donut, pretrained on the CORD dataset, to determine its capability in extracting end-to-end information from supermarket leaflet deal images. Conducted on an NVIDIA RTX 4080 (Station 1), the model has a vocabulary size of 57,580 and 201M parameters. The training configuration includes 10 epochs, validation performed every epoch, gradient clipping at 2.0, 800 training samples per epoch, a learning rate of 8e-6, batch sizes of 16 for both training and validation, 400 warmup steps, and 16-bit mixed precision.

@@

\subsubsection{Comparison of Hybrid OCR+LLM, Donut, and LVLM}

This section presents the central comparative evaluation of a hybrid OCR+LLM pipeline against end-to-end models, specifically Donut and LVLMs, for IE from supermarket deals. The assessment focuses on model performance using the best OCR+LLM configuration and two end-to-end approaches.

The evaluation framework was designed to compare the performance of three distinct model categories: OCR+LLM pipelines, end-to-end Vision Encoder-Decoder models, and LVLMs. For the OCR+LLM approach, DocTR served as the OCR engine, paired with Qwen 2.5-7B and LLaMA 3.1-8B as the large language models, selected for their comparable performance. The Donut model, an end-to-end Vision Encoder-Decoder architecture optimized for document understanding, was included to assess its capabilities in handling structured document data. Additionally, MiniCPM-V \cite{yao2024} and LLaMA 3.2-Vision \cite{touvron2023} were chosen as representative LVLMs to evaluate their multimodal reasoning abilities.

@@

\subsection{Database}
\begin{figure*}[h!]
    \centering
    \includegraphics[width=0.8\linewidth]{figures/application/db_uml.png}
    \caption{Database schema for storing metadata for supermarket leaflets and extracted deals.}
    \label{fig:application_db_uml}
\end{figure*}
For storing the extracted deals, a lightweight and efficient solution was required. SQLite was chosen due to its simplicity and fast performance, making it well-suited for handling the structured but relatively small-scale dataset. After several weeks of scraping supermarket leaflets, the dataset grew to include 540 leaflets from 20 different supermarkets. The database is structured into three primary tables (cf. \figref{fig:application_db_uml}): one for supermarket metadata, one for storing the leaflets and one containing the extracted deals. The total number of extracted deals in the database adds up to 47,829. Among these, the vast majority (46,869) are classified as standard deals, while 607 fall under product category deals, 144 correspond to coupons and 208 belong to miscellaneous categories. Since the structure and content of non-standard deals often differ significantly from regular deals featuring loyalty points, bulk discounts, or percentage-based reductions across entire product groups, the primary focus remains on standard deals. By structuring the data efficiently, the database serves as a robust backend for the subsequent web application, facilitating fast queries and scalable deal retrieval.

@@

Building on the database, a web-based application was developed to provide users with an interactive way to explore supermarket deals, which is shown in \figref{fig:application_web_overview}. The core functionality centers around an intuitive search system that allows users to find deals based on specific product names. Additionally the interface supports filtering options, enabling users to refine searches by supermarket chain or deal validity period. Each deal is displayed with its corresponding image snippet, extracted metadata, and pricing information, making it easy to compare offers across different retailers.

@@

\subsection{Implementation Details}
\label{app:implementation_details}

The experiments were conducted on two distinct computational workstations - station 1 equipped with an NVIDIA RTX 4080 GPU and station 2 with an NVIDIA GTX 1080 GPU. Therefore, the experiments could be conducted in parallel, ensuring efficient model training and evaluation.

The software stack includegraphics Python with PyTorch\footnote{\href{https://pytorch.org/}{https://pytorch.org/}} as the primary deep learning framework, HuggingFace Transformers \footnote{\href{https://huggingface.co/transformers/}{https://huggingface.co/transformers/}}
for leveraging pre-trained models, ollama \footnote{\href{https://ollama.com/}{https://ollama.com/}}
and lmstudio \footnote{\href{https://lmstudio.ai/}{https://lmstudio.ai/}} for LVLMs, and OpenCV \footnote{\href{https://opencv.org/}{https://opencv.org/}} for image preprocessing and data augmentation, as well as various other libraries for data pre-and post-processing, visualization, data management and monitoring \footnote{The codebase is available at \href{https://github.com/ilyii/leaflets}{https://github.com/ilyii/leaflets}}.
